---
layout: page
title: blah
output:
  github_document:
    pandoc_args: --webtex
  #  md_document:
#    variant: markdown_github
  preserve_yaml: true
---

```{r , echo = FALSE, eval = TRUE}
#knitr::opts_knit$set(base.dir = "/home/rafal/L2/L2projects/LegalProbabilismBNs", base.url = "https://rfl-urbaniak.github.io/LegalProbabilismBNs/")
#knitr::opts_chunk$set(fig.path = "images/")
```


### Set-up

The examples are given in R code, which we intertwine with additional explanations, which the reader is welcome to skip if they look familiar. 

First, you need to install the relevant R libraries. This is a bit tricky, because some of them have to be installed through BiocManager. One way to go is this:


\footnotesize
```{r, eval = FALSE }
install.packages("https://www.bnlearn.com/releases/bnlearn_latest.tar.gz", repos = NULL, type = "source")
install.packages("BiocManager")
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install()
BiocManager::install(c("graph", "Rgraphviz"))

install.packages("https://www.bnlearn.com/releases/bnlearn_latest.tar.gz", repos = NULL, type = "source")

```
\normalsize

Then load the libraries we use (you need to have them installed first):

```{r , options, message = FALSE, warning = FALSE}
library(bnlearn)
library(Rgraphviz)
library(knitr)
library(kableExtra)
```



While Bayes's Theorem is of immense use when it comes to calculating various conditional probabilities, if we're interested in the interaction of multiple hypotheses at various levels and multiple pieces of evidence, calculations quickly become inconvenient, to say the least. Moreover, if such considerations are to be presented to a fact-finder, it is rather unlikely that they would be transparent and easily understood. Luckily, a tool exist to make such tasks easier. Bayesian networks (BNs) can be used for a fairly transparent and computationally more manageable evalation and presentation of the interaction of multiple pieces of evidence and hypotheses. We'll start with a general presentation of BNs, and then go over a few main methods of employing BNs in presentation, aggregation and evaluation of evidence in legal fact-finding.  





```{r cancer,echo=TRUE,eval=TRUE, fig.width=2, fig.height=2, fig.align = "center",cache=TRUE}
cancer1 <- empty.graph(nodes = c("PS","SH","S","C"))
cancer1.arcs <- matrix(c("PS", "SH",
                   "PS", "S",
                   "SH", "C",
                    "S", "C"),
                 byrow = TRUE, ncol = 2,
                 dimnames = list(NULL, c("from", "to")))
arcs(cancer1) = cancer1.arcs
graphviz.plot(cancer1)
```


This can be achieved in a somewhat simpler manner like this:

```{r cancerSimpler, fig.width=2, fig.height=2, fig.align = "center",cache=TRUE}
cancer2DAG <- model2network("[SH|PS][PS][S|PS][C|SH:S]")
graphviz.plot(cancer2DAG)
```




